---
title: "The Cluster Depth Tests"
subtitle: "Massively Univariate Tests for EEG"
author: "Jaromil Frossard"
institute: "University of Geneva"
date: "2021/11/14 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---


```{r xaringanExtra, echo=FALSE}
xaringanExtra::use_tile_view()
xaringanExtra::use_clipboard()
xaringanExtra::use_scribble()
xaringanExtra::use_fit_screen()
xaringanExtra::use_progress_bar(color = "#0051BA", location = "top")
```



```{r library, include=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
library(permuco)
library(patchwork)
library(glue)
library(stringr)


library(permuco)
library(Matrix)
library(ggrepel)
library(ggpubr)

source("function/theme_jf.R")


source("script/plot_erps.R",echo = F)
source("script/plot_bonferroni.R",echo = F)
source("script/plot_perm.R",echo = F)
source("script/plot_maxt.R",echo = F)
source("script/plot_cluster.R",echo = F)
source("script/plot_cluster_hist.R",echo = F)
source("script/plot_problem_clustermass.R",echo = F)
source("script/plot_clusterdepth.R",echo = F)


```


## Table of Contents



---
class: center, middle

.pull-left[
# Testing in EEG
]




```{r erps, echo=FALSE, fig.height=6, fig.width=10, warning=FALSE, dev="svg"}
gg_erps
```

---
class: left, top

### Multiple Tests

.center[
600 ms at 512Hz = 614 Time-Points!
]

--

#### Several approaches:

- Pre-selecting a few tests (prior knowledge, miss effects)
- Average over a time frame (prior knowledge, interpretation on the time window)
- Testing all time-points (Large FWER) `r emo::ji("-1")`
- Testing all time-points + multiple comparisons procedure. `r emo::ji("+1")`

--

#### Multiple comparisons procedure:

- Parametric: Bonferroni, Holm
- Point-wise non-parametric: min-p/max-T (Westfall and Young 1993) , Troendle (Troendle 1995)
- Cluster-wise non-parametric: Cluster Mass Test (Maris and Oostenveld 2007), TFCE (Smith and Nichols 2009)
- Hybrid non-parametric: Cluster Depth Tests (Frossard and Renaud 2021)


---
class: left, top

### Family-Wise Error Rate

#### Definition
- $\text{FWER} = \text{Prob}(\text{At least 1 type 1 Error})$.
- Probability of reporting at least one false positive.

When we observe signals of length $m$, some time points are under the **null hypothesis** the other are under the **alternative hypothesis**.

Declaring significant 1 time point of a null hypothesis constitutes a type 1 error.

--

|       |non-significant|significant|        |
|:------|:-------------:|:---------:|:------:|
|True H0| U             |V          | m0     |
|True HA| T             |S          | m - m0 |
|       |m-R            |R          |m       |


$$\text{FWER} = \text{Prob}(V>0)$$


---
class: left, top

#### Example: 
- Running 10'000 experiments where **only random noise** is observed. 
- 500 of them contain at least 1 significant time point using the method $\mathcal{M}$. 
- $\mathcal{M}$ has a FWER of $\text{FWER}=500/10000 = 5\%$.

--

#### Bonferroni

- We change the univariate $\alpha$'s to $\alpha/m$.
- We multiply the univariate p-values by $m$.
- `r emo::ji("+1")` We only need univariate p-values.
- `r emo::ji("-1")` Low power.


---

class: center, top

### Univariate Tests under H0

.center[
```{r bonf, echo=FALSE, fig.height=6, fig.width=10, warning=FALSE, dev="svg"}
gg_bonf
```


]

---
class: center, middle, inverse


### Multiple comparison procedures based on resampling 


---
class: left, top

### Max-T/ min-p

\begin{align}
\text{FWER} &= \text{Prob}_{\text{H}_0}\left(\text{At least one statistic } T_i > \text{critical } T^*\right)\\
&= \text{Prob}_{\text{H}_0}\left(\text{max}_i(T_i) > \text{critical } T^*\right)
\end{align}

.center[
Same rational with p-values using the minimum!
]

--

### Distribution of the maximum

- Depend on the correlation between the tests.
- A parametric model of the correlation may not be accurate.
- Permutation tests capture the correlation under H0.

---
class: left, top


```{r perm, echo=FALSE, fig.height=8, fig.width=10, warning=FALSE, dev="svg"}
gg_perm
```


---
class: left, top


```{r perm_hist, echo=FALSE, fig.height=8, fig.width=10, warning=FALSE, dev="svg"}
gg_hist_maxt
```


---
class: left, middle

### Pluses and minuses

- `r emo::ji("+1")` Point-wise interpretation.
- `r emo::ji("+1")` Take into account the correlation.
- `r emo::ji("+1")` More powerful than parametric methods.
- `r emo::ji("-1")` Fail when there is to manly equivalent uncorrelated tests.
- `r emo::ji("-1")` Do not take into account the spatio-temporal distribution of the tests.
- `r emo::ji("-1")` For min-p, when the number of test is too large, we need to increase the number of permutations.

### Interpretation


.center[
We take an $\alpha_{\text{FWER}}$ level risk when we concluding: "All significant time points are under H1".

We will declare falsely at least 1 significant time point every $\alpha_{\text{FWER}}$ experiments containing time points under H0.
]

---
class: left, middle

### Step-wise argument

1. We sort the observed statistics $T_{(1)}>\dots>T_{(i)}>\dots> T_{(m)}$.
2. We use the multiple comparison procedure $\mathcal{M}_{1..m}$ for the decision of the hypothesis $H_{(1)}$ associated to $T_{(1)}$.
3. If we cannot reject $H_{(1)}$, then we cannot reject $H_{(1)},\dots,H_{(m)}$
4. If we reject $H_{(1)}$, then 2 cases is possible. We correctly reject $H_{(1)}$, the decision procedure $\mathcal{M}_{1..m}$ can be updated to $\mathcal{M}_{2..m}$ for the next hypothesis. We falsely rejected $H_{(1)}$, we commit a type 1 error, we can choose the procedure we want for the next hypothesis $H_{(2)},\dots,H_{(m)}$ as it will not change the FWER.


--

### Step-wise procedure

- Bonferroni + Step-wise => Holm
- min-p + Step-wise => Troendle

---
class: center, middle, inverse


# Cluster-based resampling procedure: the cluster mass test

---
class: left, top


```{r cluster, echo=FALSE, fig.height=8, fig.width=10, warning=FALSE, dev="svg"}
gg_cl
```


---
class: left, top


```{r hist_cluster, echo=FALSE, fig.height=8, fig.width=10, warning=FALSE, dev="svg"}
gg_hist_cm
```


---
class: left, top

### Pluses and minuses

- `r emo::ji("+1")` Take into account the spatio-temporal position of the tests.
- `r emo::ji("+1")` Very powerful.
- `r emo::ji("-1")` True discoveries include many false positives.
- `r emo::ji("-1")` Cluster-wise interpretation
- `r emo::ji("-1")` Weak control of the FWER.

--
### Interpretation


.center[
We take an $\alpha_{\text{FWER}}$ level risk when concluding: "There is at least 1 time point under H1".

When **all time points** are under H0, we have a $\alpha_{\text{FWER}}$ probability to observed a significant cluster.
]

---
class: left, top

```{r problem, echo=FALSE, fig.height=8, fig.width=10, warning=FALSE, dev="svg"}
gg_fp
```



---
class: center, middle, inverse


# The cluster depth tests: a point-wise interpretation 


---
class: left, top


## Rational


- From the cluster mass test, we *keep the threshold* defining clusters.
- We reject the idea of a cluster mass. We *do not aggregate the statistics*.


- The cluster depth is the position of a time point *relatively to the border of the cluster*.
- we define cluster depth "from the head" and "from the tail".

--

.center[
We test relatively to the cluster-depth: "The first time point a cluster is under H0." 

H0: *No difference at the 1st cluster depth* from the head.

Then, we generalized to all cluster-depths.
]


---
class: left, top

```{r cdepth, echo=FALSE, fig.height=8, fig.width=10, warning=FALSE, dev="svg"}
gg_cdepth
```


---
class: left, top

# References

- Frossard, J., & Renaud, O. (2021). The Cluster Depth Tests: Toward Point-Wise Strong Control of the Family-Wise Error Rate in Massively Univariate Tests with Application to M/EEG. *arXiv preprint arXiv:2105.07514*.
- Maris, E., & Oostenveld, R. (2007). Nonparametric statistical testing of EEG-and MEG-data. *Journal of neuroscience methods, 164*(1), 177-190.
- Smith, S. M., & Nichols, T. E. (2009). Threshold-free cluster enhancement: addressing problems of smoothing, threshold dependence and localisation in cluster inference. *Neuroimage, 44*(1), 83-98.
- Tipura, E., Renaud, O., & Pegna, A. J. (2019). Attention shifting and subliminal cueing under high attentional load: an EEG study using emotional faces. *Neuroreport, 30*(18), 1251-1255.
- Troendle, J. F. (1995). A stepwise resampling method of multiple hypothesis testing. *Journal of the American Statistical Association, 90*(429), 370-378.
- Westfall, P. H., & Young, S. S. (1993). *Resampling-based multiple testing: Examples and methods for p-value adjustment* (Vol. 279). John Wiley & Sons.



